Illustrating Model Properties

### 2024-05-21 manuscript.qmd
add section on midext prevalence

>>>>

{{< embed model.qmd#fig-model-prevalences-midext >}}

Lastly, in @fig-model-prevalences-midext, we plot the prevalence of midline extension in the data versus our model's prediction. It is obvious the model cannot match the large spread between early and advanced T-category seen in the data. This is because to achieve that, it would need to increase the advanced T-category patient's prior distribution over diagnosis times and at the same time reduce the probability of the tumor to cross the midline during a time-step. But since the time-priors parameter is also coupled with the spread probabilities among the LNLs, the model does not have that freedom.

<!-- Should the paragraphs below be part of the discussion? -->

However, we do not consider this discrepancy a major limitation of the model: We will not realistically be interested in the probability of midline extension, as it is always possible to assess it with high certainty. That is also the reason why we initially modelled the midline extension *not* as a random variable, but as a global risk factor that would have been turned on or off from the onset of a patient's disease evolution. This, however, lead to overly high risks for contralateral involvement in advanced T-category patients with midline extension, because then the model assumes an increased spread to the contralateral side from the onset of the disease. Which is probably not true in a majority of those cases. Thus, treating it as a random variable that only becomes true during a patient's disease evolution resulted in a better description of the data.

Formally, the wrong prediction of midline extension prevalence makes little difference, since it is always given: Instead of $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)$, we typically compute $P\left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)$, which does not suffer from the wrong probability of midline extension, as the distribution over hidden states is renormalized:

$$
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c}, \epsilon \right)}
$$

Note that a distribution over $\epsilon$ appears both in the enumerator and the denominator, which largely cancel each other, leaving only the midline extension's effect on the distribution over hidden states in the prediction.

### 2024-05-21 model.qmd
add section on midext prevalence

```{python}
#| label: fig-model-state-dist
#| fig-cap: "3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck."
from matplotlib.colors import LinearSegmentedColormap
from mpl_toolkits.axes_grid1 import AxesGrid

@@

```{python}
#| label: fig-model-prevalences-overall
#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension."
from matplotlib import gridspec
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw
from scripts import paths, shared

@@

```{python}
#| label: fig-model-prevalences-with-ipsi
#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown are four scenarios, all including contralateral LNL II involvement: Early T-category and an ipsilateral N0 neck (green), early T-category and ipsilateral LNL II involvement (blue), as well as the same two scenarios but for advanced T-category (orange and red). The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement. Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough."
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw
from scripts import paths, shared

@@

>>>>

```{python}
#| label: fig-model-state-dist
#| fig-cap: "3D state distribution of a midline model with 3 LNLs (II, III, and IV) in both sides of the neck."
from matplotlib.colors import LinearSegmentedColormap
from mpl_toolkits.axes_grid1 import AxesGrid
from lyscripts.plot.utils import COLORS

@@

```{python}
#| label: fig-model-prevalences-overall
#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown for the contralateral LNLs II (blue), III (orange), and IV (green). The top row shows scenarios with early T-category tumors, the bottom row for late T-category ones. The left column depicts scenarios where the primary tumor is clearly lateralized, the right column scenarios of tumors extending over the mid-sagittal line. This figure illustrates the model's ability to describe the prevalence of different combinations of scenarios involving the risk factors T-category and midline extension."
from matplotlib import gridspec
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS
from scripts import paths, shared

@@

```{python}
#| label: fig-model-prevalences-with-ipsi
#| fig-cap: "Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown are four scenarios, all including contralateral LNL II involvement: Early T-category and an ipsilateral N0 neck (green), early T-category and ipsilateral LNL II involvement (blue), as well as the same two scenarios but for advanced T-category (orange and red). The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement. Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough."
import matplotlib.pyplot as plt
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS
from scripts import paths, shared

@@

## Portion of Midline Extension

Since we model the tumor's midline extension as a random variable we can predict its portion among early or late T-category patients and compare it with the observed prevalence in the data.

```{python}
#| label: fig-model-prevalences-midext
#| fig-cap: "Comparing the predicted (histograms) and observed (lines depicting beta posteriors) prevalence of midline extension for early (blue) and late (orange) T-category. While the prevalence is predicted correctly when marginalizing over T-category, the model cannot capture the degree of separation observed in the data. Since the tumor's midline extension is virtually always part of the diagnosis and hence *given* when predicting a patient's risk, we do not consider this discrepancy a major issue."
import matplotlib.pyplot as plt
from lyscripts.scenario import Scenario
from lyscripts.plot.utils import Histogram, BetaPosterior, draw, COLORS
from scripts import paths, shared

fig, ax = plt.subplots()
content = []
colors = iter([COLORS[c] for c in ["blue", "orange"]])

for t_stage in ["early", "late"]:
  color = next(colors)
  content.append(
    Histogram.from_hdf5(
      filename=paths.model_dir / "full" / "prevalences_midext.hdf5",
      dataname=t_stage,
      color=color,
      label="prediction"
    )
  )
  beta_post = BetaPosterior.from_hdf5(
    filename=paths.model_dir / "full" / "prevalences_midext.hdf5",
    dataname=t_stage,
    color=color,
  )
  beta_post.kwargs["label"] = f"data: {beta_post.num_success} of {beta_post.num_total}"
  content.append(beta_post)

draw(ax, content, xlims=(0., 55.))
ax.set_title("Observed vs Predicted Prevalence\nof Midline Extension")
ax.set_yticks([])
ax.set_ylim(top=0.4)
ax.set_xlabel("prevalence [%]")

shared.split_legends(
  ax=ax,
  titles=["early T-cat.", "late T-cat."],
  locs=[(0.05, 0.99), (0.6, 0.99)],
)
